---
title: "NOAA Weather Analysis"
author: "Mark Davey"
date: "July 24, 2016"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.width=10, fig.height=10)
```

##Synopsis
Large rare individual events (such as Hurricane Katrina, the F5 Tornado spike in 2011 or the Chicago heat wave of 1995) outweigh usual events dominate this data. The weather is unpredictable, the fatalities, injuries and property damage are not evenly spread across time or location but tied to extreme events.

Surprisingly excessive heat is the leading cause of death by a wide margin and occurred over multiple years. This is a preventable cause of death given the availability of Air Conditioning but it is also has the greatest trend downward during data provided, which indicates government intervention was effective. 
Injuries are tied to tornadoes, flood and Wind (wind constant cause).

Property damage was dominated by Hurricane Katrina. Crop damage was not related to a single event (except for an ice storm in 1992) but droughts and floods.





##Data Processing
Check if the raw data file (i.e. the original .csv.bz2 file) exists if not download it.
Read in the compressed file directly but remove unneeded columns.
Make sure to use data.table since this is a large data set.
```{r message=FALSE,warning=FALSE}
#clean up
#setwd("C:/Users/mdavey/Google Drive/Data Science/reproducable research/Week 4")
rm(list=ls())
#list.files(path="./data")
#
#
library(lubridate)
library(data.table)
library(ggmap)
library(maps)
library(mapdata)
library(dplyr)
library(sqldf)
library(ggplot2)
library(reshape2)


###Download the data and read it into the R system
#Check if the files exist and download it (if not, unpack), then read it in to memory


#get the data from the remote system and unpack it only if does not exist
if (!file.exists("data")) {
  dir.create("data")
}
if (!file.exists("./data/StormData.csv.bz2")) {
  fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(fileURL,destfile = "./data/StormData.csv.bz2",method="libcurl")
}
#read the storm data into memory since this is large amount of data I'm using data.table rather than data.frame
#StormDataAll <- data.table(read.csv("./data/StormData.csv.bz2"))
requiredFields <- c("BGN_DATE","EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP","LATITUDE","LONGITUDE","REFNUM")
StormData <- data.table(subset(read.csv("./data/StormData.csv.bz2"),select=requiredFields))

```


##Data Transformation
Does the analysis include description and justification for any data transformations?

 * Fix error in data for Napa valley. Remarks and internet search show data in million not billion. I'm not going to fix all data but this was throwing off my analysis.
 * Rename events that could tag multiple items such as "HURRICANE/TYPHOON" "STORM SURGE".
 * Since property damage and crop damage have different magnitudes based upon the DMGEXP. Create a look up table with a magnitude in integer and then calculate the real value so we can sort and compare lines.
 * There is a serious problem with inconsistent event type names in this data. There are 985 unique descriptions of EVTYPE which makes analysis impossible. Create a tag dictionary so we can map keywords to 25 of the most common types. This will create a one to many relationship with certain events since for example Katrina Hurricane has "STORM SURGE" is two different tags but maps to two different categories STORM and SURGE. I can longer compare data globally except by event type.  
 * Added year from the beginning date. The documentation identifies that multiple data sources were used to create this data set. But they all had a different beginning date for instance Tornado data starts in the 1950's while Flood data does not start until the 1993. Therefore the analysis will be needing a date to start from since including Tornado data from pre 1990 will schew the results. Therefore all data in this document will be post 1990. 
 * The latitude and longitude was specified as a five or six digit number. The correct format should be a real two or three digit number  so this had to be fixed.


```{r message=FALSE,warning=FALSE}
#Serious data error Napa Valley was missed labeled in billion should be millions
StormData[StormData$REFNUM=="605943",PROPDMGEXP:="M"]
#Storm Surge
StormData[StormData$EVTYPE=="STORM SURGE",EVTYPE:="SURGE"]
#Eliminate double counting of HURRICANE & TYPHOON
StormData[StormData$EVTYPE=="HURRICANE/TYPHOON",EVTYPE:="HURRICANE"]
StormData[StormData$EVTYPE=="TROPICAL DEPRESSION",EVTYPE:="HURRICANE"]
exps <- unique(toupper(c(unique(as.character(StormData$PROPDMGEXP)),unique(as.character(StormData$CROPDMGEXP)))))
#StormDataAll[StormDataAll$CROPDMGEXP %in% c("+","?","H","-"),] #used this and found this to outlyers treated as expontial 1
#create dictionary mapping power of 10 exponetal 
expDictionary <- data.table(exp=c("K","","M","+","0","5","4","7","?","B","1","2","8","H","-","3","6"),
                            exp10=c(3,1,6,1,0,5,4,7,1,9,1,2,8,1,1,3,6))
setkey(expDictionary,exp)
setkey(StormData,CROPDMGEXP)
StormData <- StormData[expDictionary, nomatch=0,, allow.cartesian=TRUE]
setnames(StormData,"exp10","CROPDMGEXP1")
setkey(StormData,PROPDMGEXP)
StormData <- StormData[expDictionary, nomatch=0,, allow.cartesian=TRUE]
setnames(StormData,"exp10","PROPDMGEXP1")
StormData[,cropdamage := CROPDMG * (10 ** CROPDMGEXP1)]
StormData[,propdamage := PROPDMG * (10 ** PROPDMGEXP1)]

StormData[, c("date","time","fluff") := tstrsplit(as.character.POSIXt(BGN_DATE)," ", fixed=TRUE)]
StormData[, c("month","day","year") := tstrsplit(as.character.POSIXt(date),"/", fixed=TRUE)]
StormData[,year := as.numeric(year)]
StormSummaryYear <- sqldf("select year,sum(FATALITIES) as FATALITIES,sum(INJURIES) as INJURIES,sum(cropdamage) as cropdamage, sum(propdamage) as propdamage from StormData where year > 1990 group by year order by year")
StormSummaryYear


#OK Event types are a mess there are nearly 900 unique types
length(unique(StormData$EVTYPE))
#EventLets create tags for the event types so 
StormData$EVTYPE <- toupper(StormData$EVTYPE)
EventTypes <- data.table(EVTYPE=unique(StormData$EVTYPE))
#Create a tag dictionary to apply to event type so we can simplify this down to.
keyCategorises <- data.table(
  CATGKEY=c("TSTM WIND","EXCESSIVE HEAT","FOG","FLOOD","FLOOOD","WIND","WND","SNOW","BLIZZARD","HOT","HEAT","RECORD HIGH","TORNADO","TORNDAO",
            "HURRICANE","TYPHOON","RAIN","STORMXXX","HAIL","LIGHT","CLOUD","FUNNEL","COLD","FREEZ","COOL","LOW TEMP",
            "TIDE","DRY","PRECIPITATION","WARM","WET","ICE","FROST","VOLCANIC","SURF","SURGE","SEA","SWELLS","WAVE",
            "TSUNAMI","SUMMARY","FIRE","SMOKE","ACCIDENT","LANDSLIDE","HYPERTHERMIA","DOWNBURST","RIP CURRENT",
            "GUSTNADO","SLIDE","HIGH TEMPERATURE RECORD","HEAVY SHOWERS","SEVERE TURBULENCE","WATERSPOUT","EROSION",
            "REMNANTS OF FLOYD","DROUGHT","LANDSPOUT","DRIEST","TROPICAL DEPRESSION","RECORD TEMPERATURES","HIGH WATER",
            "RECORD LOW" ,"AVALANCHE","MARINE MISHAP","APACHE COUNTY","HEAVY MIX","DAM FAILURE","SOUTHEAST","BEACH EROSIN","HYPOTHERMIA",
            "SLEET","DUST","HEAVY PRECIPATATION","URBAN/SMALL","HIGHXXX","WATER SPOUT","MICROBURST","URBAN AND SMALLXXX","GLAZE",
            "WINTRY MIX","WINTER WEATHER","WAYTERSPOUT","LIGNTNING" ,"SMALL STREAM AND XXX",     
            "AVALANCE","OTHERX","ICY ROADS","RAPIDLY RISING WATER","TSTMW" ,               
            "?","WINTER MIX","MILD PATTERN","HEAVY SHOWER",         
            "URBAN SMALL","SMALL STREAM","URBAN/SML STREAM FLD","TEMPERATURE RECORD","LANDSLUMP",            
            "RECORD TEMPERATURE","MIXED PRECIP","NO SEVERE WEATHER","SML STREAM FLD","NONE",          
            "DAM BREAK","URBAN/SML STREAM FLDG","WINTERY MIX","SEICHE","TSTM",                
            "VOG","MONTHLY TEMPERATURE","WINTER WEATHER MIX","RED FLAG CRITERIA","WINTER WEATHER/MIX" ,"DROWNING" ),
  CATG=c("WIND","HEAT","FOG","FLOOD","FLOOD","WIND","WIND","SNOW","SNOW","HEAT","HEAT","HEAT","TORNADO","TORNADO","HURRICANE",
         "HURRICANE","RAIN","STORM","HAIL","LIGHTNING","CLOUD","CLOUD","COLD","COLD","COLD","COLD","TIDE","DROUGHT",
         "RAIN","HEAT","RAIN","ICE","ICE","VOLCANIC","SURGE","SURGE","SURGE","SURGE","SURGE","TSUNAMI","MISC","FIRE",
         "FIRE","ACCIDENT","LANDSLIDE","COLD","RAIN","SURGE" ,"TORNADO","LANDSLIDE","HEAT","RAIN","WIND","TORNADO",
         "LANDSLIDE","HURRICANE","DROUGHT","TORNADO","DROUGHT","HURRICANE","HEAT","FLOOD","COLD","AVALANCHE","ACCIDENT","MISC",
         "RAIN","ACCIDENT","MISC","MISC","COLD","SNOW","DUST","RAIN","MISC","MISC","MISC","RAIN","MISC","ICE",
         "SNOW","SNOW","TORNADO","LIGHTNING" ,"MISC",     
         "AVALANCHE","MISC","ICE","FLOOD","MISC","MISC","SNOW","MISC","RAIN",         
         "MISC","MISC","FLOOD","HEAT","LANDSLIDE","HEAT","RAIN","MISC","FLOOD","MISC",          
         "ACCIDENT","FLOOD","SNOW","SURGE","WIND","FOG","MISC","SNOW","MISC","SNOW" ,"ACCIDENT" ))
#How many unique tags for EventTypes
length(unique(keyCategorises$CATG))
#confirm no untagged eventtypes
EventTypesRemain <- data.frame(EVTTYPE=EventTypes[!grepl(paste(keyCategorises$CATGKEY,collapse="|"),EventTypes$EVTTYPE),] )
#Use a little bit of sql to create a look up table of eventtypes to tags Note this is a 1->M relationship
#since some event types such as "HOT and DROUGHT" map to two tags "HOT","DROUGHT" but this is needed since the evaluation by tag
GeneralEventTypes <- 
  data.table(sqldf("select distinct et.EVTYPE,kc.CATG,kc.CATGKEY from EventTypes et inner join keyCategorises kc on instr(et.EVTYPE,kc.CATGKEY) > 0"))
setkey(GeneralEventTypes,EVTYPE)
setkey(StormData,EVTYPE)
StormData <- StormData[GeneralEventTypes, nomatch=0,, allow.cartesian=TRUE]

```

## Data Analysis

Normalizing the four categories being analyzed so it can be viewed in a single figure. This should useful to analyze correlations between these columns. 

```{r}
StormSummaryCatYear <- sqldf("select CATG as Event_Type,year,sum(FATALITIES)/100 as [FATALITIES per 100],sum(INJURIES)/1000 as [INJURIES per 1000],sum(cropdamage)/(1e+09) as [Crop Damage in Billions Dollars], sum(propdamage)/(1e+010) as [Property Damage in Hundred Billions Dollars] from StormData where year > 1990 group by CATG,year order by CATG,year")
StormSummaryCatYear <- sqldf("select CATG as Event_Type,year,sum(FATALITIES)/100 as [FATALITIES per 100],sum(INJURIES)/1000 as [INJURIES per 1000],sum(cropdamage)/(1e+09) as [Crop Damage in Billions Dollars], sum(propdamage)/(1e+010) as [Property Damage in Hundred Billions Dollars] from StormData where year > 1990 and CATG in ('DROUGHT','FLOOD','HURRICANE','STORM','ICE','WIND','TORNADO','SURGE','HEAT','LIGHTNING') group by CATG,year order by CATG,year")
mx<-melt(StormSummaryCatYear,id.vars = c("Event_Type","year"))
plot6 <- ggplot(data=mx,aes(x=year,y=value,group=variable,color=Event_Type))+geom_line()+geom_point()+facet_grid(.~variable)+labs(x="Year",y="See Sub Title",title="NOAA data summary by category type and year")
plot6

StormSummary <- sqldf("select CATG,sum(FATALITIES) as FATALITIES,sum(INJURIES) as INJURIES,sum(cropdamage) as Crop_Damage, sum(propdamage) as Property_Damage,count(*) as [# of Events] from StormData where year > 1990 group by CATG")

```


#Results
This defines natural disasters tend to be tied to Hurriances and Tornadoes, while droughts and wind damage are a constant situation.
Significant weather events create outlying data points which distort any underlying data. The period of greater than 50 years is outside the period of the effective data obtained so other event types may not have been measured. 

##Which types of events are most harmful to population health?

###Fatalities 
```{r}
head(StormSummary[order(StormSummary$FATALITIES,decreasing=TRUE),],10)
```
 * Heat is the leading cause of death and  particular in the Midwest (Especially Chicago) experienced two events that caused. There has been a significant decrease in deaths since 2000 with a decreasing trend. The US government has pushed Cooling Assistance along with designated cooling locations during heat waves. 
 * Surprisingly wind is the second leading cause of deaths mainly due to Thunderstorms, but no specific cause or event with the fatalities spread by both region and date. 
 * The third leading cause of death is Tornado's but again a single event the F5 cluster in 2011 accounted for approximately 30% of these fatalities.
 * Flood and Surf deaths also heavily relate to a single incident  

###Injuries
```{r}
head(StormSummary[order(StormSummary$INJURIES,decreasing=TRUE),],10)
```

Tornado is the leading cause of injuries. Wind is second leading cause of injuries but it is a constant problem wide spread.
Notable events that caused the most injuries was F5 cluster in 2011 and Flooding due to Hurricane Madeline in 1998.

##Which types of events have the greatest economic consequences?

###Crop Damage (Economic)
```{r}

head(StormSummary[order(StormSummary$Crop_Damage,decreasing=TRUE),],10)
```

Drought is the leading cause of crop loss with  flooding a close second. The problem with drought is an ongoing problem which appears every few years in different regions of the country. Flooding is also a very common cause of crop damage. The remaining key events Hurricane & Ice are tied to unusual natural events that occur infrequently.

###Property Damage (Economic)
```{r}
head(StormSummary[order(StormSummary$Property_Damage,decreasing=TRUE),],10)
```

The majority of property damage is related to hurricane and Storm surge from Hurricane Katrina. There was also a cluster of hurricanes that also cause a lot of damage in 2004. Flooding is the leading causing of damage. 

##Trends by Event Type 

During the period which can be analyzed, was a trend identifiable. Plotting each major category by time to identify possible changes. 

```{r}
StormSummaryCatYear <- sqldf("select CATG as Event_Type,year,sum(FATALITIES)/100 as [FATALITIES per 100],sum(INJURIES)/1000 as [INJURIES per 1000],sum(cropdamage)/(1e+09) as [Crop Damage in Billions Dollars], sum(propdamage)/(1e+010) as [Property Damage in Hundred Billions Dollars] from StormData where year > 1990 and CATG in ('DROUGHT','FLOOD','HURRICANE','STORM','ICE','WIND','TORNADO','SURGE','HEAT','LIGHTNING') group by CATG,year order by CATG,year")
mx<-melt(StormSummaryCatYear,id.vars = c("Event_Type","year"))

 plot6 <- ggplot(data=mx,aes(x=year,y=value,group=Event_Type,color=variable))+
   geom_line()+geom_point()+
   facet_wrap(~ Event_Type)+labs(x="Year",y="See Sub Title",title="NOAA Data Summary by event type and year") + 
   coord_cartesian(ylim=c(0,5)) +  geom_smooth(method='lm')
 plot6

```

Using linear modelling, it was interesting to see if any type of event had a specific trend. The only event type that had a specific trend was heat related deaths which showed a definite decrease. All other trend where dominated by a few large events or showed no trend. 

##Location for fatalities 

Determine if specific locations have increased likelihood of weather related fatalities. 

```{r}
fatal_by_loc<- sqldf("select CATG,LATITUDE*0.01 as LATITUDE,LONGITUDE*(-0.01) as LONGITUDE,FATALITIES from StormData where LATITUDE>2513 and LONGITUDE>6500 and LONGITUDE<12470 and FATALITIES>0 and year > 1990")
states <- map_data("state")
 ggplot(data = states) + 
   geom_polygon(aes(x = long, y = lat, group = group), color = "white") + 
   geom_point(data=fatal_by_loc, aes(x=LONGITUDE,y=LATITUDE,color=CATG,size=FATALITIES)) +
   coord_fixed(1.3) + labs(title="NOAA Reported Fatalities by Locaton and Type") +
  guides(fill=FALSE)
```   

Certain event types do not have a location such as Excessive Heat 
Location is more specific for certain types of events. 

 * Tornado's cluster around the Midwest and Midsouth.

 * Flooding is highly prevalent across Texas and to a lesser extent across the entire east half of the U.S.

 * Reviewing a population density map and correlating the above map 
   * West Coast and East Coast population density do not match the fatalities on this map. 
   * The Far West is under populated so lack of fatalities may be due to this.
   * Texas and South are prone to weather issues.
